---
title: "Project 1"
author: "Emily Surrell"
date: 10/5/2026
format: html
execute:
  error: true
---

Bring in libraries needed for this project.

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
library(hms)
library(quantreg)
```

### Introduction

For this project, we will be working with 2021, 2022, 2023, and 2024 data from the Public Use Microdata Sample (PUMS) Census API. First, we obtain an API key from [the Census API website](https://api.census.gov/data/key_signup.html).

```{r}
#save API key into a variable that can easily be used again later
key <- "6ba7e209537814e77a7b7fc5aa8df5552ef8591b"
```

Sample API call to show that the key works:

```{r}
exampleAPI <- GET(paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=",
  key))

#view outline of results
names(exampleAPI)

#view status code
exampleAPI$status_code
```

A status code of 200 means that the API call was successful.

### Part 1: Data Processing

#### Parsing Variable Values

First, we will practice parsing variable values using 2021 variables.

```{r}
#Call the API:
vars2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

#Parse the JSON:
vars2021 <- fromJSON(rawToChar(vars2021$content))

#View example:
vars2021$variables$AGEP$values

#View example:
vars2021$variables$SEX$values

#Collect variable values into a list:
vars2021 <- list(
  AGEP = vars2021$variables$AGEP$values,
  GASP = vars2021$variables$GASP$values,
  GRPIP = vars2021$variables$GRPIP$values,
  JWAP = vars2021$variables$JWAP$values,
  JWDP = vars2021$variables$JWDP$values,
  JWMNP = vars2021$variables$JWMNP$values,
  PWGTP = vars2021$variables$PWGTP$values,
  FER = vars2021$variables$FER$values,
  HHL = vars2021$variables$HHL$values,
  SCH = vars2021$variables$SCH$values,
  SCHL = vars2021$variables$SCHL$values,
  SEX = vars2021$variables$SEX$values,
  STATE = vars2021$variables$ST$values,
  REGION = vars2021$variables$REGION$values,
  DIVISION = vars2021$variables$DIVISION$values
  )

#view results
names(vars2021)
```

That worked great for the 2021 data. Now let's generalize this into a function that can be used for different years:

```{r}
get_var_values <- function(year){
  
  #Validate inputs
  if (!(year %in% c(2021, 2022, 2023, 2024))){
    stop("Invalid year: Please enter 2021, 2022, 2023, or 2024")
  }

  #API call
  vars <- GET(paste0("https://api.census.gov/data/", 
                   year, 
                   "/acs/acs1/pums/variables.json"))

  #Parse the JSON
  vars <- fromJSON(rawToChar(vars$content))

  #Collect variable values into a list
  if (year %in% c(2021, 2022)){
    vars <- list(
      AGEP = vars$variables$AGEP$values,
      GASP = vars$variables$GASP$values,
      GRPIP = vars$variables$GRPIP$values,
      JWAP = vars$variables$JWAP$values,
      JWDP = vars$variables$JWDP$values,
      JWMNP = vars$variables$JWMNP$values,
      PWGTP = vars$variables$PWGTP$values,
      FER = vars$variables$FER$values,
      HHL = vars$variables$HHL$values,
      SCH = vars$variables$SCH$values,
      SCHL = vars$variables$SCHL$values,
      SEX = vars$variables$SEX$values,
      STATE = vars$variables$ST$values,
        #NOTE: ST is used instead of STATE in 2021 and 2022
      REGION = vars$variables$REGION$values,
      DIVISION = vars$variables$DIVISION$values
      )
  } else if (year %in% c(2023, 2024)){
    vars <- list(
      AGEP = vars$variables$AGEP$values,
      GASP = vars$variables$GASP$values,
      GRPIP = vars$variables$GRPIP$values,
      JWAP = vars$variables$JWAP$values,
      JWDP = vars$variables$JWDP$values,
      JWMNP = vars$variables$JWMNP$values,
      PWGTP = vars$variables$PWGTP$values,
      FER = vars$variables$FER$values,
      HHL = vars$variables$HHL$values,
      SCH = vars$variables$SCH$values,
      SCHL = vars$variables$SCHL$values,
      SEX = vars$variables$SEX$values,
      STATE = vars$variables$STATE$values,
        #NOTE: STATE is used instead of ST in 2023 and 2024
      REGION = vars$variables$REGION$values,
      DIVISION = vars$variables$DIVISION$values
      )
  }
  
  #Return the list of variables
  return(vars)
}
```

Test to show that the function checks for an invalid year:

```{r}
get_var_values(2025)
```

Use the function to get the variable values in each of the four years.

```{r}
vars2021 <- get_var_values(2021)
vars2022 <- get_var_values(2022)
vars2023 <- get_var_values(2023)
vars2024 <- get_var_values(2024)
```

Combine the four years into a single `list` object (a `list` of four `list`s).

```{r}
allvars <- list(vars2021 = vars2021,
                vars2022 = vars2022,
                vars2023 = vars2023,
                vars2024 = vars2024)

#View results
names(allvars)

#View 2021 variables
names(allvars$vars2021)

#View REGION variable from 2021
allvars$vars2021$REGION$item
```

Write the combined list object to a `.rds` file.

```{r}
write_rds(allvars, "data/variables.rds")
```

#### Parsing API Data

Begin with an example of a single API call:

```{r}
pums2024 <- GET(paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums?get=", #year
  "PWGTP,AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,", #numeric variables
  "FER,HHL,SCH,SCHL,SEX", #categorical variables
  "&ucgid=0400000US37&key=", #geography
  key))
```

Parse the data returned by the `GET()` function:

```{r}
content <- fromJSON(rawToChar(pums2024$content))

#look at results
head(content)

#look at class
class(content)

#look at type
typeof(content)

#look at structure
str(content)
```

What we got was a character matrix, where the first row contains the headers. Set the column names to be these headers.

```{r}
colnames(content) <- content[1,]

#look at results
head(content)
```
Then, drop the first row with the headers that we no longer need.

```{r}
content <- content[-1,]

#look at results
head(content)
```

Format as a `tibble`.

```{r}
astibble <- as_tibble(content)

#look at results
astibble
```

Everything is formatted as `<chr>`. Let's make all variables numeric.

```{r}
astibble <- astibble |>
  mutate(across(everything(), as.numeric))

#look at results
astibble
```

That is a pretty good start. We will need to do further cleaning, but let's generalize the process up to this point into a helper function to take what is returned by `GET()` and turn it into a tibble that is full of `<dbl>` values:

```{r}
parse_from_get <- function(getobject){
  
  #parse data:
  content <- fromJSON(rawToChar(getobject$content))
  
  #assign column names:
  colnames(content) <- content[1,]
  
  #remove header row:
  content <- content[-1,]

  #format as numeric tibble:
  tibble <- as_tibble(content) |>
    mutate(across(everything(), as.numeric))
  
  #return tibble:
  return(tibble)
}
```

Example:

```{r}
parse_from_get(pums2024)
```

Great! Now continue with cleaning. We have a tibble with proper column names, but everything is stored as `<dbl>`.

Turn categorical variables into factors using variable levels from the `.rds` file. First, prepare the variable levels. Using SCH as an example:

```{r}
#read from .rds file
rds <- read_rds("data/variables.rds")

#narrow down to where the keys and values are
sch2024_values <- rds$vars2024$SCH$item

#view example:
names(sch2024_values)

#view example:
sch2024_values

#begin a data.frame to store keys and values
sch2024_clean <- data.frame(rownum = 1:length(as.numeric(names(sch2024_values))))

#view example:
sch2024_clean

#loop through keys and values and assign them to the data.frame
for (i in 1:length(as.numeric(names(sch2024_values)))){
  #assign key
  sch2024_clean$num[i] <- as.numeric(names(sch2024_values))[i]
  #assign value
  sch2024_clean$value[i] <- sch2024_values[[i]]
}

#view example:
sch2024_clean
```

Next, update the tibble to apply the key-value pairs for our factor variable SCH.

```{r}
astibble <- astibble |>
  mutate(SCH = factor(SCH, sch2024_clean$num, sch2024_clean$value))

#view results
astibble |>
  select("SCH")
```

Now SCH is a factor variable!

But, there are many variables that we may need to convert into factor variables.

Generalize this process into a tidy-style function that translates a numeric variable into a factor variable. It will take a `tibble` as its first parameter, and return a `tibble`.

```{r}
convert_to_factor <- function(tibble, year, variable){
  
  #validate inputs
  if (!(is_tibble(tibble))) {
    stop("First argument is not a tibble.")
  } else if (!(year %in% c(2021, 2022, 2023, 2024))){
    stop("Invalid year: Please enter 2021, 2022, 2023, or 2024")
  }
  
  #read in the .rds file
  rds <- read_rds("data/variables.rds")
  
  #narrow down to where the specific variable's levels are
  values <- rds[[paste0("vars", year)]]
  values <- values[[variable]]
  values <- values$item
  
  #initialize a data.frame to hold the levels
  cleaned <- data.frame(rownum = 1:length(as.numeric(names(values))))

  #loop through keys and levels and assign them to the data.frame
  for (i in 1:length(as.numeric(names(values)))){
    #assign keys
    cleaned$num[i] <- as.numeric(names(values))[i]
    #assign values
    cleaned$value[i] <- values[[i]]
  }

  #update the tibble to format the variable as a factor!
  tibble[[variable]] <- factor(tibble[[variable]],
                      cleaned$num, 
                      cleaned$value)
  
  #return the updated tibble
  return(tibble)
}
```

With this function, we should now be able to convert the other variables to factors very easily.

```{r}
astibble <- astibble |>
  convert_to_factor(2024, "FER") |>
  convert_to_factor(2024, "HHL") |>
  convert_to_factor(2024, "SCHL") |>
  convert_to_factor(2024, "SEX") |>
  convert_to_factor(2024, "STATE")

#view results
astibble |>
  select(FER, HHL, SCH, SCHL, SEX, STATE)
```

Now all the categorical variables are stored as factors!

Next, clean the time variables. First, we can convert them to factors and then to characters:

```{r}
astibble <- astibble |>
  convert_to_factor(2024, "JWAP") |>
  convert_to_factor(2024, "JWDP") |>
  mutate(across(c(JWAP, JWDP), as.character))

#view results
astibble |>
  select(JWAP, JWDP)
```

Two examples of values in these columns would be "7:30 a.m. to 7:34 a.m." and "6:10 a.m. to 6:14 a.m.". Another example is "N/A (not a worker; worker who worked from home)", for which we will want to assign an `NA` value. Store these example values in a vector.

```{r}
strings <- c("7:30 a.m. to 7:34 a.m.",
             "6:10 a.m. to 6:14 a.m.",
             "N/A (not a worker; worker who worked from home)")

#view vector
strings
```

Use the `word()` function to split the strings into a starting time and an ending time.

```{r}
start <- word(strings, 1, sep = " to ")
end <- word(strings, -1, sep = " to ")

#view starting times and ending times split apart
start
end
```

Now that the starting time and the ending time are separated, convert each of them from character format to a time format.

```{r}
start <- as_hms(parse_date_time(start, orders ="HM", quiet = TRUE))
end <- as_hms(parse_date_time(end, orders = "HM", quiet = TRUE))

#view results
start
end
```

In order to find the midpoint, we need to format these values into a numeric matrix.

```{r}
time_example <- matrix(c(as.numeric(start), as.numeric(end)), ncol=2)

#view results
time_example
```

Find the midpoint between the starting time and the ending time.

```{r}
midpoint <- rowMeans(time_example)

#view results
midpoint
```

Lastly, format the midpoint back into a time format.

```{r}
midpoint <- as_hms(midpoint)

#view results
midpoint
```

Now formulate this process into a tidy-style function!

```{r}
time_midpoint <- function(tibble, timecolumn){
  
  #validate inputs
  if (!(is_tibble(tibble))) {
    stop("First argument is not a tibble.")
  } else if (!(timecolumn %in% c("JWAP", "JWDP"))){
    stop('Invalid timecolumn: Please enter "JWAP" or "JWDP"')
  }
  
  #clean timecolumn
  
  tibble[[timecolumn]] <- as_hms(rowMeans(matrix(c(
  
  as.numeric(as_hms(parse_date_time(
    word(tibble[[timecolumn]], 1, sep = " to "), 
    orders ="HM",
    quiet = TRUE))),
  
  as.numeric(as_hms(parse_date_time(
    word(tibble[[timecolumn]], -1, sep = " to "), 
    orders ="HM",
    quiet = TRUE)))),
  
  ncol = 2)))
  
  return(tibble)
}

```

Now test this on our example tibble:

```{r}
astibble <- astibble |>
  time_midpoint("JWAP") |>
  time_midpoint("JWDP")

#view results
astibble |>
  select(JWAP, JWDP)
```

Test to show input validation:

```{r}
astibble |>
  time_midpoint("AGEP")
```

Add **census** class:

```{r}
class(astibble) <- c("census", class(astibble))

#view class:
class(astibble)
```

#### Combining It All Together

Now that we have gone through a full example of cleaning a single API call, we can use the helper functions we created to create an overall function which queries the API for a single year and then does all the data cleaning.

```{r}
query <- function(year = 2024,
                  nvar = c("PWGTP", "AGEP"),
                  cvar = "SEX",
                  geolevel = "STATE",
                  geovalue = NULL){
  
  #Validate geolevel input and translate into numeric geolevel:
  if (geolevel == "STATE"){
    geolevel_translated <- "040"
    
    #set default geovalue for state
    if (is.null(geovalue)){
      geovalue <- 37 #North Carolina
    }
  } else if (geolevel == "REGION"){
    geolevel_translated <- "020"
    
    #set default geovalue for region
    if (is.null(geovalue)){
      geovalue <- 3 #South
    }
  } else if (geolevel == "DIVISION"){
    geolevel_translated <- "030"
    
    #set default geovalue for division
    if (is.null(geovalue)){
      geovalue <- 5 #South Atlantic
    }
  } else {
    stop('Please enter "STATE", "REGION", or "DIVISION" for geolevel.')
  }
  
  #More input validation:
  if (!(year %in% c(2021, 2022, 2023, 2024))){
    stop("Invalid year: please enter 2021, 2022, 2023, or 2024")
  } else if (!("PWGTP" %in% nvar)){
    stop('Please include "PWGTP" in list of numeric variables.')
  } else if (length(nvar) < 2){
    stop('Please include at least one other numeric varible besides "PWGTP".')
  } else if (length(cvar) < 1){
    stop("Please include at least one categorical variable.")
  }
  
  #Validate numeric variables:
  for (i in nvar){
    if (!(i %in% c("PWGTP","AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP"))){
      stop('Invalid numeric variable. Please enter "PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", or "JWMNP".')
    }
  }
  
  #Validate character variables:
  for (i in cvar){
    if (!(i %in% c("FER","HHL","SCH","SCHL","SEX"))){
      stop('Invalid categorical varibale. Please enter "FER", "HHL", "SCH", "SCHL", or "SEX".')
    }
  }
  
  #Formulate geolevel and geovalue portion of API call:
  if (geovalue == "all"){
    geo_api <- paste0(",", geolevel)
    if (year %in% c(2021,2022) & geolevel == "STATE"){
      geo_api <- ",ST"
    }
  } else {
    geo_api <- paste0("&ucgid=", geolevel_translated,
    "0000US", geovalue)
  }
  
  #API call:
  raw_api <- GET(paste0("https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    paste(nvar, collapse = ","),
    ",",
    paste(cvar, collapse = ","),
    geo_api,
    "&key=",
    key))

  #Parse and convert into a numeric tibble using helper function:
  tibble <- parse_from_get(raw_api)
  
  #For 2021 and 2022: rename ST variable to STATE:
  if (year %in% c(2021, 2022) & geolevel == "STATE"){
    tibble <- tibble |>
      rename("STATE" = "ST")
  }

  #Convert character variables and geolevel to factors using helper function:
  for (i in c(cvar, geolevel)){
    tibble <- tibble |>
      convert_to_factor(year, i)
  }
  
  #Convert time variables using helper functions:
  if ("JWAP" %in% nvar){
    tibble <- tibble |>
      convert_to_factor(year, "JWAP") |>
      time_midpoint("JWAP")
  }
  if ("JWDP" %in% nvar){
    tibble <- tibble |>
      convert_to_factor(year, "JWDP") |>
      time_midpoint("JWDP")
  }
  
  #Add census class:
  class(tibble) <- c("census", class(tibble))
  
  return(tibble)
}
```

Examples using the function:

```{r}
example1 <- query(2024, c("PWGTP", "AGEP"), "SEX", "STATE", 37)

#view results
example1

example2 <- query(2023, c("PWGTP", "AGEP", "GASP"), c("SEX", "FER"), "REGION", 1)

#view results
example2

example3 <- query(2021, c("PWGTP", "JWAP", "JWDP"), "SCH", "STATE", 36)

#view results
example3

example4 <- query(2024)

#view results
example4

example5 <- query(2022, c("PWGTP", "AGEP"), "SEX", "DIVISION", "all")

#view results
example5
```

Tests to show input validation:

```{r}
#Examples showing invalid inputs:
query(2025)
query(2024, c("JWAP"))
query(2024, c("PWGTP", "JWAP"), c())
query(2023, c("PWGTP", "JWAP"), "SCH", "AREA")
```

#### Combining Years

Write a function that calls the single-year function as many times as needed and then combines the data into one final tibble with a year variable.

```{r}
multiyear_query <- function(years, nvar, cvar, geolevel, geovalue){
  
  #initialize a flag to keep track of if it is the first year
  firstyear <- 1
  
  for (year in years){
    
    if(firstyear==1){
      
      #for the first year: run the single-year query and assign the year variable
      tibble <- query(year, nvar, cvar, geolevel, geovalue) |>
        mutate(year = as.character(year))
      #note: formatting year as character for categorical analysis
  
      #change the firstyearflag for the next iteration of the for loop
      firstyear <- 0
    } else {
      
      #for every year after the first year, append new rows to the working tibble
      tibble <- dplyr::bind_rows(tibble,
      query(year, nvar, cvar, geolevel, geovalue) |>
        mutate(year = as.character(year)))
    }
  }
  
  #after all years have been processed, return the final tibble
  return(tibble)
}
```

Examples:

```{r}
example6 <- multiyear_query(c(2023, 2024), c("PWGTP", "AGEP"), "SEX", "REGION", "all")

#view results
example6

example7 <- multiyear_query(c(2021, 2023), c("PWGTP", "JWAP"), c("FER", "HHL"), "DIVISION", 9)

#view results
example7

#all North Carolina data:
example8 <- multiyear_query(2021:2024,
    c("PWGTP", "AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP"),
    c("FER", "HHL", "SCH", "SCHL", "SEX"),
    "STATE", 37)

#view results
example8
```

### Part 2: Writing a Generic Function for Summarizing

#### Numeric Summaries

Write a function that produces means and standard deviations for our numeric variable(s) (ignore the time variables) and counts for our categorical variable(s) (year should be treated as categorical).

```{r}
summary.census <- function(tibble,
                           nvar = NULL,
                           cvar = NULL){
  
  #set defaults if variables are not specified
  if (is.null(nvar)){
    #default is all numeric variables except for PWGTP and the time variables
    nvar <- attributes(tibble)$names
    nvar <- nvar[nvar %in% c("AGEP","GASP","GRPIP","JWMNP")]
  }
   
  if (is.null(cvar)){ 
    #default is all categorical variables
    cvar <- attributes(tibble)$names
    cvar <- cvar[cvar %in% c("FER","HHL","SCH","SCHL","SEX","STATE","REGION","DIVISION","year")]
  }
  
  #initialize empty lists for the results
  numeric_results <- list()
  categorical_results <- list()
  
  #summarize numeric variables
  for (x in nvar){
    sample_mean <- sum(tibble[[x]]*tibble$PWGTP) / sum(tibble$PWGTP)
    
    sample_sd <- sqrt(sum(tibble[[x]]^2*tibble$PWGTP) / sum(tibble$PWGTP)-sample_mean^2)
    
    numeric_results[[x]] <- c(sample_mean = sample_mean,
                         sample_sd = sample_sd)
  }
  
  #summarize categorical variables
  for (x in cvar){
    categorical_results[[x]] <- tibble[,c("PWGTP", x)] |>
      group_by(across(-PWGTP)) |>
      summarize(count = sum(PWGTP))
  }
  
  results <- list(numeric = numeric_results,
                  categorical = categorical_results)
  
  return(results)
}
```

Examples:

```{r}
summary(example8, c("AGEP", "GASP"), c("SEX", "SCH", "year"))
summary(example8)
```

#### Graphical Summaries

Create a generic `plot()` function for a `census` class tibble which creates a boxplot of a numeric variable at the levels of the categorical variable.

```{r}
plot.census <- function(tibble, num_var, cat_var){

#initialize the plot
ggplot(tibble) +
    
    #add boxplot layer
    geom_boxplot(aes(x = get(cat_var),
                     y = get(num_var),
                     fill = get(cat_var),
                     weight = PWGTP)) +
    
    #set x and y labels and title
    labs(title = paste(num_var, cat_var, sep = " by "),
         x = cat_var,
         y = num_var) +
    
    #set legend label
    scale_fill_discrete(cat_var) +
    
    #set the color theme
    theme_light() +
  
    #center the title
    theme(plot.title = element_text(hjust = 0.5)) +
    
    #angle the x-axis labels
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

Examples:

```{r}
plot(example8, "AGEP", "SEX")
plot(example8, "GASP", "year")
```

### Investigation

Investigate something interesting from the data using the functions above!

For this investigation, we will look at the numeric variable GPIP, which represents "Gross rent as a percentage of household income past 12 months".

First, query the data.

```{r}
rent_data <- multiyear_query(2021:2024, c("PWGTP", "GRPIP"), c("SEX", "SCHL"), "STATE", 37)

#view results
rent_data
```

Look at a summary of the data.

```{r}
summary(rent_data)

#Summarize year specifically:
summary(rent_data, cvar = "year")

#Summarize GRPIP and SCHL specifically:
summary(rent_data, "GRPIP", "SCHL")
```

See how the rent-to-income ratio is the same or different across sex, educational attainment, and year.

```{r}
plot(rent_data, "GRPIP", "SEX")
plot(rent_data |>
       filter(SCHL %in% c("Regular high school diploma",
                          "Associate's degree",
                          "Bachelor's degree",
                          "Master's degree",
                          "Doctorate degree")), 
     "GRPIP", "SCHL")
plot(rent_data, "GRPIP", "year")
```

