---
title: "Project 1"
author: "Emily Surrell"
date: 9/30/2026
format: html
---

Bring in libraries needed for this project.

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
```

This project works with 2021, 2022, 2023, and 2024 data from the Public Use Microdata Sample (PUMS) Census API. First, obtain an API key from [the Census API website](https://api.census.gov/data/key_signup.html).

```{r}
key <- "6ba7e209537814e77a7b7fc5aa8df5552ef8591b"
```

Sample API call to show that the key works:

```{r}
pums2024 <- GET(paste0("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=", key))
```

### Part 1: Data Processing

#### Parsing Variable Values

Get 2021 variable values:

```{r}

#Call the API:
vars2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

#Parse the JSON:
vars2021 <- fromJSON(rawToChar(vars2021$content))

#Collect variable values into a list:
vars2021 <- list(
  AGEP = vars2021$variables$AGEP$values,
  GASP = vars2021$variables$GASP$values,
  GRPIP = vars2021$variables$GRPIP$values,
  JWAP = vars2021$variables$JWAP$values,
  JWDP = vars2021$variables$JWDP$values,
  JWMNP = vars2021$variables$JWMNP$values,
  PWGTP = vars2021$variables$PWGTP$values,
  FER = vars2021$variables$FER$values,
  HHL = vars2021$variables$HHL$values,
  SCH = vars2021$variables$SCH$values,
  SCHL = vars2021$variables$SCHL$values,
  SEX = vars2021$variables$SEX$values,
  ST = vars2021$variables$ST$values,
  REGION = vars2021$variables$REGION$values,
  DIVISION = vars2021$variables$DIVISION$values
  )
```

Make this into a function that can be used for different years:

```{r}

get_var_values <- function(year){

  vars <- GET(paste0("https://api.census.gov/data/", 
                   year, 
                   "/acs/acs1/pums/variables.json"))

  vars <- fromJSON(rawToChar(vars$content))

  if (year == 2021 | year == 2022){
    vars <- list(
      AGEP = vars$variables$AGEP$values,
      GASP = vars$variables$GASP$values,
      GRPIP = vars$variables$GRPIP$values,
      JWAP = vars$variables$JWAP$values,
      JWDP = vars$variables$JWDP$values,
      JWMNP = vars$variables$JWMNP$values,
      PWGTP = vars$variables$PWGTP$values,
      FER = vars$variables$FER$values,
      HHL = vars$variables$HHL$values,
      SCH = vars$variables$SCH$values,
      SCHL = vars$variables$SCHL$values,
      SEX = vars$variables$SEX$values,
      ST = vars$variables$ST$values,
      REGION = vars$variables$REGION$values,
      DIVISION = vars$variables$DIVISION$values
      )
  }
  
  else if (year == 2023 | year == 2024){
    vars <- list(
      AGEP = vars$variables$AGEP$values,
      GASP = vars$variables$GASP$values,
      GRPIP = vars$variables$GRPIP$values,
      JWAP = vars$variables$JWAP$values,
      JWDP = vars$variables$JWDP$values,
      JWMNP = vars$variables$JWMNP$values,
      PWGTP = vars$variables$PWGTP$values,
      FER = vars$variables$FER$values,
      HHL = vars$variables$HHL$values,
      SCH = vars$variables$SCH$values,
      SCHL = vars$variables$SCHL$values,
      SEX = vars$variables$SEX$values,
      STATE = vars$variables$STATE$values, #STATE is used instead of ST
      REGION = vars$variables$REGION$values,
      DIVISION = vars$variables$DIVISION$values
      )
    }
  
  return(vars)
}
```

Use the function to get the variable values in each of the four years.

```{r}
vars2021 <- get_var_values(2021)
vars2022 <- get_var_values(2022)
vars2023 <- get_var_values(2023)
vars2024 <- get_var_values(2024)
```

Combine into a single list object.

```{r}
allvars <- list(vars2021 = vars2021,
                vars2022 = vars2022,
                vars2023 = vars2023,
                vars2024 = vars2024)
```

Write to a `.rds` file.

```{r}
write_rds(allvars, "data/variables.rds")
```

