---
title: "Project 1"
author: "Emily Surrell"
date: 9/30/2026
format: html
---

Bring in libraries needed for this project.

```{r}
library(httr)
library(jsonlite)
library(tidyverse)
library(dplyr)
library(lubridate)
```

### Introduction

For this project, we will be working with 2021, 2022, 2023, and 2024 data from the Public Use Microdata Sample (PUMS) Census API. First, we obtain an API key from [the Census API website](https://api.census.gov/data/key_signup.html).

```{r}
#save API key into a variable that can easily be used again later
key <- "6ba7e209537814e77a7b7fc5aa8df5552ef8591b"
```

Sample API call to show that the key works:

```{r}
exampleAPI <- GET(paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24&key=",
  key))

#view results
names(exampleAPI)

exampleAPI$status_code
```

A status code of 200 means that the API call was successful.

### Part 1: Data Processing

#### Parsing Variable Values

First, we will practice parsing variable values using 2021 variables.

```{r}

#Call the API:
vars2021 <- GET("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")

#Parse the JSON:
vars2021 <- fromJSON(rawToChar(vars2021$content))

#View examples of values:
vars2021$variables$AGEP$values
vars2021$variables$SEX$values

#Collect variable values into a list:
vars2021 <- list(
  AGEP = vars2021$variables$AGEP$values,
  GASP = vars2021$variables$GASP$values,
  GRPIP = vars2021$variables$GRPIP$values,
  JWAP = vars2021$variables$JWAP$values,
  JWDP = vars2021$variables$JWDP$values,
  JWMNP = vars2021$variables$JWMNP$values,
  PWGTP = vars2021$variables$PWGTP$values,
  FER = vars2021$variables$FER$values,
  HHL = vars2021$variables$HHL$values,
  SCH = vars2021$variables$SCH$values,
  SCHL = vars2021$variables$SCHL$values,
  SEX = vars2021$variables$SEX$values,
  ST = vars2021$variables$ST$values,
  REGION = vars2021$variables$REGION$values,
  DIVISION = vars2021$variables$DIVISION$values
  )

#view results
names(vars2021)
```

That worked great for the 2021 data. Now let's generalize this into a function that can be used for different years:

```{r}

get_var_values <- function(year){

  #API call
  vars <- GET(paste0("https://api.census.gov/data/", 
                   year, 
                   "/acs/acs1/pums/variables.json"))

  #Parse the JSON
  vars <- fromJSON(rawToChar(vars$content))

  #Collect variable values into a list
  if (year == 2021 | year == 2022){
    vars <- list(
      AGEP = vars$variables$AGEP$values,
      GASP = vars$variables$GASP$values,
      GRPIP = vars$variables$GRPIP$values,
      JWAP = vars$variables$JWAP$values,
      JWDP = vars$variables$JWDP$values,
      JWMNP = vars$variables$JWMNP$values,
      PWGTP = vars$variables$PWGTP$values,
      FER = vars$variables$FER$values,
      HHL = vars$variables$HHL$values,
      SCH = vars$variables$SCH$values,
      SCHL = vars$variables$SCHL$values,
      SEX = vars$variables$SEX$values,
      ST = vars$variables$ST$values,
      REGION = vars$variables$REGION$values,
      DIVISION = vars$variables$DIVISION$values
      )
  }
  
  else if (year == 2023 | year == 2024){
    vars <- list(
      AGEP = vars$variables$AGEP$values,
      GASP = vars$variables$GASP$values,
      GRPIP = vars$variables$GRPIP$values,
      JWAP = vars$variables$JWAP$values,
      JWDP = vars$variables$JWDP$values,
      JWMNP = vars$variables$JWMNP$values,
      PWGTP = vars$variables$PWGTP$values,
      FER = vars$variables$FER$values,
      HHL = vars$variables$HHL$values,
      SCH = vars$variables$SCH$values,
      SCHL = vars$variables$SCHL$values,
      SEX = vars$variables$SEX$values,
      STATE = vars$variables$STATE$values,
#NOTE: STATE is used instead of ST in 2023 and 2024
      REGION = vars$variables$REGION$values,
      DIVISION = vars$variables$DIVISION$values
      )
    }
  
  #Return the list of variables
  return(vars)
}
```

Use the function to get the variable values in each of the four years.

```{r}
vars2021 <- get_var_values(2021)
vars2022 <- get_var_values(2022)
vars2023 <- get_var_values(2023)
vars2024 <- get_var_values(2024)
```

Combine the four years into a single list object (a list of four lists).

```{r}
allvars <- list(vars2021 = vars2021,
                vars2022 = vars2022,
                vars2023 = vars2023,
                vars2024 = vars2024)

#view results
names(allvars)
names(allvars$vars2021)
allvars$vars2021$REGION$item
```

Write the combined list object to a `.rds` file.

```{r}
write_rds(allvars, "data/variables.rds")
```

#### Parsing API Data

Begin with an example of a single API call:

```{r}
pums2024 <- GET(paste0(
  "https://api.census.gov/data/2024/acs/acs1/pums?get=", #year
  "PWGTP,AGEP,GASP,GRPIP,JWAP,JWDP,JWMNP,", #numeric variables
  "FER,HHL,SCH,SCHL,SEX", #categorical variables
  "&ucgid=0400000US37&key=", #geography
  key))
```

Parse the data returned by the `GET()` function:

```{r}
content <- fromJSON(rawToChar(pums2024$content))

#look at results
head(content)
```

Remove the first row, which contains headers.

```{r}
content <- content[-1,]

#look at results
head(content)
```

Transform all the values from character into numeric.

```{r}
content <- matrix(as.numeric(content), ncol = ncol(content))

#look at results
head(content)
```

Format as a `tibble`.

```{r}
astibble <- as_tibble(content)

#look at results
astibble
```

That is a pretty good start. We will need to fix the column names and do further cleaning, but let's generalize the process up to this point into a helper function to take what is returned by `GET()` and turn it into a numeric tibble:

```{r}
parse_from_get <- function(getobject){
  
  #parse data:
  content <- fromJSON(rawToChar(getobject$content))
  
  #remove header row:
  content <- content[-1,]
  
  #make numeric:
  content <- matrix(as.numeric(content), ncol = ncol(content))

  #format as tibble and return:
  return(as_tibble(content))
}
```

Example:

```{r}
parse_from_get(pums2024)
```

Great! Now continue with cleaning, starting with fixing the column names.

```{r}
colnames(astibble) <- c("PWGTP","AGEP","GASP","GRPIP","JWAP","JWDP","JWMNP",
                        "FER","HHL","SCH","SCHL","SEX",
                        "STATE")

#view results
astibble
```

Now we have a tibble with proper column names, but everything is stored as `<dbl>`.

Turn categorical variables into factors using variable levels from the `.rds` file. First, prepare the variable levels. Using SCH as an example:

```{r}
#read from .rds file
rds <- read_rds("data/variables.rds")

#narrow down to where the keys and values are
sch2024_values <- rds$vars2024$SCH$item

#begin a data.frame to store keys and values
sch2024_clean <- data.frame(rownum = 1:length(as.numeric(names(sch2024_values))))

#loop through keys and values and assign them to the data.frame
for (i in 1:length(as.numeric(names(sch2024_values)))){
  #assign key
  sch2024_clean$num[i] <- as.numeric(names(sch2024_values))[i]
  #assign value
  sch2024_clean$value[i] <- sch2024_values[[i]]
}

#view the cleaned data.frame which holds the levels
sch2024_clean
```

Next, update the tibble to apply the key-value pairs for our factor variable SCH.

```{r}
astibble <- astibble |>
  mutate(SCH = factor(SCH, sch2024_clean$num, sch2024_clean$value))

#view results
astibble |>
  select("SCH")
```

Now SCH is a factor variable!

But, there are many variables that we may need to convert into factor variables.

Generalize this process into a tidy-style function that translates a numeric variable into a factor. It will take a `tibble` as its first parameter, and return a `tibble`.

```{r}

convert_to_factor <- function(tibble, year, variable){
  
  #read in the .rds file
  rds <- read_rds("data/variables.rds")
  
  #narrow down to where the specific variable's levels are
  values <- rds[[paste0("vars", year)]]
  values <- values[[variable]]
  values <- values$item
  
  #initalize a data.frame to hold the levels
  cleaned <- data.frame(rownum = 1:length(as.numeric(names(values))))

  #loop through keys and levels and assign them to the data.frame
  for (i in 1:length(as.numeric(names(values)))){
    #assign keys
    cleaned$num[i] <- as.numeric(names(values))[i]
    #assign values
    cleaned$value[i] <- values[[i]]
  }

  #update the tibble to format the variable as a factor!
  tibble[[variable]] <- factor(tibble[[variable]],
                      cleaned$num, 
                      cleaned$value)
  
  #return the updated tibble
  return(tibble)
}

```

With this function, we should now be able to convert the other variables to factors very easily.

```{r}
astibble <- astibble |>
  convert_to_factor(2024, "FER") |>
  convert_to_factor(2024, "HHL") |>
  convert_to_factor(2024, "SCHL") |>
  convert_to_factor(2024, "SEX") |>
  convert_to_factor(2024, "STATE")

#view results
astibble |>
  select(FER, HHL, SCH, SCHL, SEX, STATE)
```

Now all the categorical variables are stored as factors!

Next, clean the time variables. **THIS PART IS STILL IN PROGRESS**

```{r}
test <- astibble |>
  convert_to_factor(2024, "JWAP") |>
  convert_to_factor(2024, "JWDP")
```

Add **census** class:

```{r}
class(astibble) <- c("census", class(astibble))
```

#### Combining It All Together

Now that we have gone through a full example of cleaning a single API call, we can use the helper functions we created to create an overall function which queries the API for a single year.

```{r}
query <- function(year, nvar, cvar, geolevel, geovalue){
  
  #translate geolevel into numeric geolevel:
  if (geolevel == "STATE"){geolevel_translated <- "040"}
  else if (geolevel == "REGION"){geolevel_translated <- "020"}
  else if (geolevel == "DIVISION"){geolevel_translated < "030"}
  
  #API call:
  raw_api <- GET(paste0("https://api.census.gov/data/",
    year,
    "/acs/acs1/pums?get=",
    paste(nvar, collapse = ","),
    ",",
    paste(cvar, collapse = ","),
    "&ucgid=",
    geolevel_translated,
    "0000US",
    geovalue,
    "&key=",
    key))

  #Parse:
  tibble <- parse_from_get(raw_api)

  #Fix column names:
  colnames(tibble) <- c(nvar, cvar, geolevel)

  #Convert character variables and geolevel to factors:
  for (i in c(cvar, geolevel)){
    tibble <- tibble |>
      convert_to_factor(year, i)
  }
  
  #Add census class:
  class(tibble) <- c("census", class(tibble))
  
  return(tibble)
}
```

Examples using the function:

```{r}
query(2024, c("PWGTP", "AGEP"), "SEX", "STATE", 37)
query(2023, c("PWGTP", "AGEP", "GASP"), c("SEX", "FER"), "REGION", 1)
```

#### Combining Years

Write a function that calls the single-year function as many times as needed and then combines the data into one final tibble with a year variable.

```{r}
multiyear_query <- function(years, nvar, cvar, geolevel, geovalue){
  
  firstyear <- 1
  
  for (year in years){
    
    if(firstyear==1){
      tibble <- query(year, nvar, cvar, geolevel, geovalue) |>
        mutate(year = year)
  
      firstyear <- 0
    }
    else {
      tibble <- dplyr::bind_rows(tibble,
      query(year, nvar, cvar, geolevel, geovalue) |>
        mutate(year = year))
    }
  }
  
  return(tibble)
}
```

Examples:

```{r}
multiyear <- multiyear_query(c(2023, 2024), c("PWGTP", "AGEP"), "SEX", "STATE", 37)

multiyear
```

### Part 2: Writing a Generic Function for Summarizing

#### Numeric Summaries

#### Graphical Summaries

### Investigation

Investigate something interesting from the data using the functions above!
